---
title: "Sentiment Analysis for Mental Health - PREPARING DATASET FOR DistilBERT DEEP LEARNING MODEL"
output: html_document
date: "2024-11-27"
---
# Loading the original dataset
```{r}
# Set the file path for the CSV file
file_path2 <- "C:/Users/jivko/Documents/Data Analytics, Big Data, and Predictive Analytics/Personal Project/Sentiment Analysis for Mental Health/Combined Data.csv"


# Read the CSV file into a dataframe
sentiment_analysis <- read.csv(file_path2, header = TRUE)
```

# Printing the first few rows of the dataframe
```{r}
print(head(sentiment_analysis))
```
# Removing redundant X column
```{r}
sentiment_analysis_use <- sentiment_analysis[, !names(sentiment_analysis) %in% c("X")]
```

```{r}
print(head(sentiment_analysis_use))
```

# Structure of dataset
```{r}
str(sentiment_analysis_use)
```

# Check for missing values
```{r}
# Check for missing values in each column
colSums(is.na(sentiment_analysis_use))
```

# Distribution of mental health statuses
```{r}
status_counts <- table(sentiment_analysis_use$status)
print(status_counts)
```

# Matching each status count to median (3888)
```{r}
# Load the libraries
library(dplyr)
library(caret)

# Assuming your dataset is called sentiment_analysis
# Get the distribution of the classes in the sentiment_analysis dataset
class_counts <- table(sentiment_analysis_use$status)

# Initialize an empty list to hold the balanced dataset
balanced_data <- list()

# Loop through each class
for (class in names(class_counts)) {
  # Subset the data for the current class
  class_data <- sentiment_analysis_use %>% filter(status == class)
  
  # If the class has fewer than 3888 samples, oversample
  if (nrow(class_data) < 3888) {
    # Oversample with replacement
    class_data <- class_data[sample(1:nrow(class_data), 3888, replace = TRUE), ]
  }
  
  # If the class has more than 3888 samples, undersample
  else if (nrow(class_data) > 3888) {
    # Undersample to 3888 samples
    class_data <- class_data[sample(1:nrow(class_data), 3888), ]
  }
  
  # Add the balanced class data to the list
  balanced_data[[class]] <- class_data
}

# Combine the balanced data
balanced_data <- do.call(rbind, balanced_data)

# Check the distribution of the balanced data
balanced_class_counts <- table(balanced_data$status)
print(balanced_class_counts)
```
# Checking structure of balanced data
```{r}
str(balanced_data)
```
# Setting a fixed sample size per class (25% of 3888)
```{r}
# Set a fixed sample size per class (e.g., 25% of 3888)
fixed_sample_size <- 972  # Round down to ensure consistency across classes

# Perform stratified sampling
sampled_balanced_data <- do.call(rbind, lapply(split(balanced_data, balanced_data$status), function(class_data) {
  class_data[sample(1:nrow(class_data), fixed_sample_size), ]
}))

# Check the new distribution
table(sampled_balanced_data$status)
```
# Structure of sampled balanced dataset
```{r}
str(sampled_balanced_data)
```

# Light preprocessing of text data and exporting as CSV for DistilBERT to use
```{r}
# Load required libraries
library(textclean)  # For replace_contraction and text cleaning
library(tm)         # For corpus and text manipulation

# Replace stemming with lemmatization
corpus <- Corpus(VectorSource(sampled_balanced_data$statement))

# Apply minimal preprocessing steps
corpus <- tm_map(corpus, content_transformer(tolower))  # Convert to lowercase
corpus <- tm_map(corpus, stripWhitespace)                # Strip extra whitespaces

# Remove non-text symbols and replace ’ with '
corpus <- tm_map(corpus, content_transformer(function(x) gsub("[^[:alnum:] [:space:]]", "", x)))  # Remove non-alphanumeric symbols
corpus <- tm_map(corpus, content_transformer(function(x) gsub("’", "'", x)))  # Fix ’ to '


# Convert cleaned corpus to a data frame
cleaned_statements_lots <- data.frame(statement = sapply(corpus, as.character),
                                      status = sampled_balanced_data$status)

# View a sample of the cleaned data
head(cleaned_statements_lots)

# Optional: Export the cleaned data to CSV
file_path <- "C:/Users/jivko/Documents/Data Analytics, Big Data, and Predictive Analytics/Personal Project/Sentiment Analysis for Mental Health/Cleaned_Statements_try1.csv"
write.csv(cleaned_statements_lots, file = file_path, row.names = FALSE)

# Confirm that the file has been saved
cat("CSV file saved at:", file_path)
```



